---
output:
  md_document:
    variant: markdown_github
---

nanostringr
==========

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

An R Package for quality assurance checking, normalization and batch effects adjustments of NanoString data suitable for single sample processing. This is the companion R package for the paper published in XXX.

## Installation
To install this package, use devtools:
```{r}
#devtools::install_github("OVCARE/nanostringr")
```

## Dependencies
Please note that the package CHL26predictor (https://github.com/tinyheero/CHL26predictor) is needed to run the Hodgkin Lymphoma predictive model.

## Overview
To see the full list of exported functions:

```{r}
library(nanostringr)
ls("package:nanostringr")
```

A quick overview of the key functions:

- NanoStringQC: Computes quality assurance metrics.
- HKnorm: Performs log (base 2) transformation and normalization to Housekeeping Genes
- refMethod: Performs batch effect correction using the reference-based strategie 

A vignette that reproduces most of the analyses in the paper is included.
