---
title: "QAmeasures"
output: html_document
---


```{r setupQualityMeasures, message = FALSE, echo = FALSE, warning = FALSE}
library(knitr)
library(dplyr)
opts_chunk$set(message = FALSE, echo = FALSE, warning = FALSE)


# Define colors constants----
COL.OVD <- "#66C2A5"
COL.OVO <- "#A6D854"
COL.OVCL <- "#FC8D62"
COL.HLD <-  "#8DA0CB"
COL.HLO <-"#E78AC3"
```

The dataset expQC is the annotation file for all of the experiments, run in different CodeSets, including:

- Hodgkin Lymphoma Clinical Samples (HL) 	TOTAL 	74 samples	
    - HL1	n=32	Unique Samples
    - HL2	n=32	Replicates of HL1 samples
    - HL3	n=10	Replicates (subset of HL1 samples)

- Ovarian Cancer Clinical Samples (OC)		TOTAL	258	samples
    - OC1	n=129	Unique
    - OC2	n=129	Replicates of OC1 samples
	    
- Ovarian Cancer Cell Lines (OVCL)	TOTAL	26 
    - OC1	n=13	Unique
    - OC2	n=13	Replicates of OC1 samples
	
- DNA Oligonucleotides for the HL CodeSet (HLO)	TOTAL	68 HLO pool run at different concentrations
    - HL1	n=36	
    - HL2	n=30	
    - HL3	n=2	
    
- DNA Oligonucleotides for the OC CodeSet (OVO)		TOTAL	135	OVO pool run at different concentrations
    - OC1	n=47	
	  - OC2	n=88	

The NanoStringQC function computes several QC metrics and appends them to the annotation matrix. Care should be taken to ensure that the data is in the proper format.

```{r message=TRUE,echo=TRUE}
library(nanostringr)
expOVD <- NanoStringQC(ovd.r,subset(expQC,OVD=="Yes"))
expOVO <- NanoStringQC(ovo.r,subset(expQC,OVO=="Yes"))
expOVCL <- NanoStringQC(ovc.r,subset(expQC,OVCL=="Yes"))
expHLD <- NanoStringQC(hld.r,subset(expQC,HLD=="Yes"))
expHLO <- NanoStringQC(hlo.r,subset(expQC,HLO=="Yes"))
expQC <- rbind(expHLD,expOVD,expHLO,expOVO,expOVCL)
expQC$cohort <- factor(c(rep("HLD", nrow(expHLD)), 
                rep("OVD", nrow(expOVD)),rep("HLO",nrow(expHLO)),
                rep("OVO", nrow(expOVO)), rep("OVCL", nrow(expOVCL))))
expQC <- expQC %>% 
  mutate(cohort = factor(stringr::str_replace_all(cohort,
                                                   c("HLD" = "HL",
                                                     "OVD" = "OC")),
                          levels = c("HL", "OC", "OVCL", "HLO", "OVO")))

```


# Metrics for Quality Assurance


## Fields Of View (FOV) 
 

```{r perFOVPlot, fig.cap =  "Samples that failed imaging QC based on percent fields of view (FOV) counted across cohorts"}

boxplot(perFOV~cohort, ylab = "% FOV", main = "% FOV by Cohort", data = expQC, pch = 20, col = c(COL.HLD, COL.OVD, COL.OVCL,COL.HLO,COL.OVO));
abline(h = 75, lty = 2, col = "red")
grid(NULL, NULL, lwd = 1)

```


## Postive Controls
    
### Linearity of Positive Controls

```{r linPCPlot, fig.cap ="Plot of $R^2$ of postive control probles from samples across all cohorts."}
boxplot(linPC~cohort, ylab = expression(R^2), main = "Linearity of Positive Controls by Cohort", data = expQC, pch = 20, col = c(COL.HLD, COL.OVD, COL.OVCL,COL.HLO,COL.OVO), ylim=c(0,1));
abline(h = 0.95, lty = 2, col = "red")
grid(NULL, NULL, lwd = 1)
```


## Signal to Noise Ratio (SNR)

### Level of Housekeeping Genes

```{r averageHKPlot, fig.cap ="Average log expression of Housekeeping genes by Cohort."}
boxplot(averageHK~cohort, ylab = "Average log HK expression", main = "Average log expression of Housekeeping genes by Cohort", data = expQC, pch = 20, col = c(COL.HLD, COL.OVD, COL.OVCL,COL.HLO,COL.OVO));
abline(h = 50, lty = 2, col = "red")
grid(NULL, NULL, lwd = 1)
```


### Limit of Detection (LOD)

```{r lodPlot, fig.cap ="Limit of detection by cohort." }
boxplot(lod~cohort, ylab = "LOD", main = "Limit of detection (LOD) by Cohort", data = expQC, pch = 20, col = c(COL.HLD, COL.OVD, COL.OVCL,COL.HLO,COL.OVO));
abline(h = 50, lty = 2, col = "red")
grid(NULL, NULL, lwd = 1)
```


```{r pergdPlot,fig.cap ="Percent genes of total (excluding controls) detected above the limit of detection"}
boxplot(pergd~cohort, data = expQC, border = "white",
       ylab = "% Genes Detected", 
       main = "Percent of Genes Detected Above \n the Limit of Detection", 
       pch = 20, col = c(COL.HLD, COL.OVD, COL.OVCL,COL.HLO,COL.OVO));
abline(h = 50, lty = 2, col = "red")
grid(NULL, NULL, lwd = 1)
stripchart(pergd ~ cohort, data = expQC, 
           vertical = TRUE, method = "jitter",  
           pch = 20, cex = 0.4 , col = "#3A6EE3", 
           add = TRUE) 
```



```{r snPlot,fig.cap ="Signal to Noise versus % Gene Detected by cohort"}
sn <- 100
detect <- 60

plot(expOVD$sn, expOVD$pergd, pch = 20, col = COL.OVD, xaxt = "n", ylim = c(0,100), xlim = range(expOVD$sn),
     xlab = "Signal to Noise Ratio",ylab = "% Genes Detected")
points(expOVO$sn, expOVO$pergd, pch = 20, col = COL.OVO)
points(expOVCL$sn, expOVCL$pergd, pch = 20, col = COL.OVCL)
points(expHLD$sn, expHLD$pergd, pch = 20, col = COL.HLD)
points(expHLO$sn, expHLO$pergd, pch = 20, col = COL.HLO)
axis(1, at = seq(0, max(expQC$sn) + 1, 300))
abline(v = sn, col = "red", lwd = 2)
abline(h = detect, lty = 2)
title("Signal to Noise vs \n Ratio of Genes Detected")
legend("bottomright", c("HL","OC","OVCL","HLO","OVO"), pch = 20, bty = 'n', col = c(COL.HLD,COL.OVD, COL.OVCL,COL.HLO, COL.OVO))
```

```{r snZoom, fig.cap ="Signal to Noise versus % Gene Detected by cohort zoomed in to the area of possible failures"}

plot(expOVD$sn, expOVD$pergd, pch = 20, col = COL.OVD, xaxt = "n", ylim = c(0,100), xlim = c(0,6000),
     xlab = "Signal to Noise Ratio ",ylab = "Ratio of Genes Detected")
points(expOVO$sn, expOVO$pergd, pch = 20, col = COL.OVO)
points(expOVCL$sn, expOVCL$pergd, pch = 20, col = COL.OVCL)
points(expHLD$sn, expHLD$pergd, pch = 20, col = COL.HLD)
points(expHLO$sn, expHLO$pergd, pch = 20, col = COL.HLO)
axis(1, at = seq(0, max(expQC$sn) + 1, 300))

abline(v = sn, col = "red", lwd = 2)
abline(h = detect, lty = 2)
title("Signal to Noise vs \n Ratio of Genes Detected (Zooming-in)")
legend("bottomright", c("HL","OC","OVCL","HLO","OVO"), pch = 20, bty = 'n', col = c(COL.HLD,COL.OVD, COL.OVCL,COL.HLO, COL.OVO))

```


## Binding Density (BD)

```{r bdPlot, fig.cap = "Binding density by cohort. Samples outside of the dashed lines are flagged for having failed the binding density QC measure"}
boxplot(binding.density~cohort, ylab = "Binding Density", main = "Binding Density (BD) by Cohort", data = expQC, pch = 20, col = c(COL.HLD, COL.OVD, COL.OVCL,COL.HLO,COL.OVO));
abline(h = 0.05, lty = 2, col = "red")
abline(h = 2.25, lty = 2, col = "red")
grid(NULL, NULL, lwd = 1)
```


```{r lodbd, echo = FALSE}
plot(expQC$lod, type = "l", lwd = 1, col = "coral",
     ylab = "LOD", xlab = "Samples", main = "LOD and Binding Density")
points(which(expQC$bdFlag == "Failed"), expQC$lod[expQC$bdFlag == "Failed"], col = "blue", pch = 19)
grid(NA, 5, lwd = 1)
legend("topleft", legend = "Failed Binding Density QC", col = "blue", pch = 19, bty = 'n')

plot(expQC$binding.density, expQC$lod, pch = 20, col = "deepskyblue", lwd = 1, xlim = c(0,5), ylim = c(-6,60), xaxt = "n", panel.first = grid(NULL,NULL, lwd = 1), ylab = "LOD", xlab = "Binding Density", main = "LOD vs Binding Density")
axis(1, at = seq(0, 5, by = .5), las = 2)
points(expQC$binding.density[expQC$bdFlag == "Failed"], expQC$lod[expQC$bdFlag == "Failed"], col = "blue", pch = 19)
legend("topright", legend = "Failed Binding Density QC", col = "blue", pch = 20, bty = 'n')
```
